---
title: "mixed-models-workshop"
output: html_document
date: '2023-03-10'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What are mixed models and why should we use them?

**Mixed effects models ** are an extension of linear models that include a combination of fixed and random effects (Bolker et al. 2009; Harrison et al. 2018). 

Mixed effects models and the introduction of random effects have a number of key (non-mutually exclusive) benefits (adapted from Harrison et al. 2018): 

  1.  Biological datasets are often highly structured, containing clusters of non-independent observational units that are hierarchical in nature, and mixed models allow us to explicitly model the non-independence in such data. 
  2.  Explicit modelling of the random effects structure will aid correct inference about fixed effects, depending on which level of the system’s hierarchy is being manipulated. 
  3.  Random effects typically represent some grouping variable and allow the estimation of variance in the response variable within and among these groups, which reduces the probability of false positives (Type I error rates) and false negatives (Type II error rates).


## Mixed model resources: 

For more details, code, and more information on using mixed effects models see the following online resources: 

  1. [Our coding club](https://ourcodingclub.github.io/tutorials/mixed-models/>)

And the following published articles and books: 

  1. [Harrison et al. 2018 A brief introduction to mixed effects modelling and multi-model inference in ecology](<https://peerj.com/articles/4794/)
  2. [Bolker et al. 2009 Generalized linear mixed models: a practical guide for ecology and evolution](https://www.sciencedirect.com/science/article/pii/S0169534709000196?via%3Dihub)
  3. [Zuur et al. 2009 Mixed Effects Models and Extensions in Ecology with R](https://link.springer.com/book/10.1007/978-0-387-87458-6)


## load key libraries
```{r lme4}
library(Matrix)
library(lme4)
library(nlme)
library(ggplot2)
library(broom.mixed)
```

## When should we use mixed models: 



## Getting started with syntax:

Mixed model syntax using the `lmer`, `nlme`, and `glmmTMB` packages is similar. 



```{r, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
